A weakly structured stem for human origins in Africa

### 2022-08-12 paper.tex
change sigma to SE

In contrast, under the multiple-merger model, stem populations merge with
varying proportions to form the different contemporary groups. 
We observe a sharp bottleneck in Stem 1 down to $N_e=117$ after the split of
the Neanderthal branch. This represents the lower bound allowed in our
optimization (i.e. 1\% of the ancestral $N_e$), although the size of this
bottleneck is poorly constrained ($\sigma_{N_e}=838$). After a long period of
exchange with Stem 2, Stem 1 then fractures into ``Stem 1E'' and ``Stem 1S''
479ka. The timing of this divergence was also poorly constrained ($\sigma_T=
166$ka). These populations evolve independently until 119ka
($\sigma_T=
28$ka) when Stem 1S and Stem 2 combine to form the ancestors of the Nama, with
proportions 29\%, 71\% respectively. Similarly, Stem 1E and Stem 2 combine in
equal proportions (50\% each) to form the ancestors of the western Africans and
eastern Africans (and thus also all individuals who later disperse during the
Out of Africa event). Finally, the Mende receive a large additional pulse of
gene flow from Stem 2, replacing 18\% of their population 25ka ($\sigma_T=
0.6$ka).
The later Stem 2 contribution to the western African Mende resulted in
better model fits ($\Delta LL \approx 60,000$).
Speculatively, this may indicate that an ancestral Stem 2 population occupied western or
Central Africa, broadly speaking. The differing proportions in the Nama and
eastern Africans may also indicate geographic separation of Stem 1S in southern
Africa and Stem 1E in eastern Africa.

@@

A population reticulation involves multiple stems contributing genetically to
the formation of a group. One way in which this can happen is through the
geographic expansion of one or both stems. For example, if during MIS 5, either
Stem 1S (Figure~\ref{fig:best-fit-models}B) 
from southern Africa moved northward thereby encountering Stem 2,
or Stem 2 moved from central/western Africa southward into Stem 1S -- then we
could observe disproportionate ancestry contributions from different stems in
contemporary groups. We observed two merger events. The first, between Stem 1S and
Stem 2, results in the formation of an ancestral Khoe-San population $\approx120$ka.
The second $\approx100$ka between Stem 1E and Stem 2, results in the formation of the
ancestors of East/West Africans as well as later ``Out of Africans''. The rapid
rise in sea levels and increased precipitation during MIS 5e, following a
glacial period of aridity across Africa \citep{Blome2012-lw}, might have
triggered migration inland away from the coasts, as has been suggested, e.g., for the
Paleo-Agulhas plain \citep{Marean2014-pg}.

Following these merger events, the stems subsequently fracture into
subpopulations which then appear to persist over the past $\sim120$ka. These
subpopulations can be linked to contemporary groups despite subsequent gene flow
across the continent; for example, a genetic lineage sampled in the Gumuz has a
$0.44$ probability of being inherited from the ancestral ‘eastern’ subpopulation
(Stem 1E) $150$ka versus $0.03$ probability of being inherited from the ‘southern’
subpopulation (Stem 1S) and $0.53$ probability of being inherited from Stem 2
(see Table~\ref{tab:supp-lineage-probabilitys} for additional comparisons). We also
find that Stem 2 continued to contribute to western Africans during the Last
Glacial Period, indicative that this gene flow likely occurred in
western/central Africa (Table~\ref{tab:migration-rates}).

>>>>

In contrast, under the multiple-merger model, stem populations merge with
varying proportions to form the different contemporary groups. 
We observe a sharp bottleneck in Stem 1 down to $N_e=117$ after the split of
the Neanderthal branch. This represents the lower bound allowed in our
optimization (i.e. 1\% of the ancestral $N_e$), although the size of this
bottleneck is poorly constrained (standard error (SE) $=838$). After a long period of
exchange with Stem 2, Stem 1 then fractures into ``Stem 1E'' and ``Stem 1S''
479ka. The timing of this divergence was also poorly constrained (SE $=166$ka).
These populations evolve independently until 119ka
(SE $=28$ka) when Stem 1S and Stem 2 combine to form the ancestors of the Nama, with
proportions 29\%, 71\% respectively. Similarly, Stem 1E and Stem 2 combine in
equal proportions (50\% each) to form the ancestors of the western Africans and
eastern Africans (and thus also all individuals who later disperse during the
Out of Africa event). Finally, the Mende receive a large additional pulse of
gene flow from Stem 2, replacing 18\% of their population 25ka (SE $=0.6$ka).
The later Stem 2 contribution to the western African Mende resulted in
better model fits ($\Delta LL \approx 60,000$).
Speculatively, this may indicate that an ancestral Stem 2 population occupied western or
Central Africa, broadly speaking. The differing proportions in the Nama and
eastern Africans may also indicate geographic separation of Stem 1S in southern
Africa and Stem 1E in eastern Africa.

@@

A population reticulation involves multiple stems contributing genetically to
the formation of a group. One way in which this can happen is through the
geographic expansion of one or both stems. For example, if during MIS 5, either
Stem 1S (Figure~\ref{fig:best-fit-models}B) 
from southern Africa moved northward thereby encountering Stem 2,
or Stem 2 moved from central/western Africa southward into Stem 1S -- then we
could observe disproportionate ancestry contributions from different stems in
contemporary groups. We observed two merger events. The first, between Stem 1S and
Stem 2, results in the formation of an ancestral Khoe-San population $\approx120$ka.
The second event $\approx100$ka between Stem 1E and Stem 2,
results in the formation of the
ancestors of East/West Africans as well as later ``Out of Africans''. The rapid
rise in sea levels and increased precipitation during MIS 5e, following a
glacial period of aridity across Africa \citep{Blome2012-lw}, might have
triggered migration inland away from the coasts, as has been suggested, e.g., for the
Paleo-Agulhas plain \citep{Marean2014-pg}.

Following these merger events, the stems subsequently fracture into
subpopulations which then appear to persist over the past $\sim$120ka. These
subpopulations can be linked to contemporary groups despite subsequent gene flow
across the continent; for example, a genetic lineage sampled in the Gumuz has a
$0.44$ probability of being inherited from the ancestral ‘eastern’ subpopulation
(Stem 1E) $150$ka versus $0.03$ probability of being inherited from the ‘southern’
subpopulation (Stem 1S) and $0.53$ probability of being inherited from Stem 2
(see Table~\ref{tab:supp-lineage-probabilitys} for additional comparisons). We also
find that Stem 2 continued to contribute to western Africans during the Last
Glacial Period, indicative that this gene flow likely occurred in
western/central Africa (Table~\ref{tab:migration-rates}).
